---
title: 'Exercises'
output: 
  html_document:
    toc: false
    code_folding: hide
---


```{r setup, echo=FALSE, purl = FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, eval = FALSE, cache = FALSE)
options(width = 80)
```

\  

## Exercise 5 (optional): Visualising data using ggplot

\  

Alternative (optional) solutions to Exercise 5 for those who use (or are interested in using) the `ggplot` approach to plotting data. See [Chapter 5](https://intro2r.com/graphics_r.html) of the Introduction to R book for more information about how to use `ggplot`.

**Base R graphics are what is taught and what you will be assessed on.** Nothing here is required, and you can ignore this page entirely without missing anything. It is here because some of you will already know `ggplot2`, or will want to learn it, and it is worth seeing the same plots built both ways.

\  

```{r packages}

# make ggplot2, gridExtra and GGally packages available
# Note: you may need to install these packages first
# Use : install.packages('ggplot2', dep = TRUE)
# Use : install.packages('gridExtra', dep = TRUE)
# Use : install.packages('GGally', dep = TRUE)

library(ggplot2)
library(gridExtra)
library(GGally)
```

\  

4\. Import the data and recode `sex` and `smoking` as factors. This part is identical whichever plotting system you use: `ggplot2` replaces your graphics code, not your data handling.

```{r Q4}
cardiac <- read.table('data/cardiacdata.txt', header = TRUE, sep = "\t",
                      na.strings = "NA", stringsAsFactors = TRUE)

cardiac$Fsex <- factor(cardiac$sex, levels = c(1, 2),
                       labels = c("Female", "Male"))

cardiac$Fsmoking <- factor(cardiac$smoking, levels = c(1, 2, 3),
                           labels = c("Current", "Ex", "Never"))

str(cardiac)
```

\  

5\. Patients in each combination of smoking status and sex. A table is still the right tool for this, but if you want it as a plot, `geom_bar()` will count for you.

```{r Q5}
table(cardiac$Fsmoking, cardiac$Fsex)

ggplot(data = cardiac, aes(x = Fsmoking, fill = Fsex)) +
  geom_bar(position = "dodge") +
  labs(x = "Smoking status", y = "Number of patients", fill = "Sex")
```

\  

6\. Cleveland dotplots for `bmi`, `systolic`, `tchol` and `alcohol`. There is no `geom_dotchart()`; you build one by plotting the value against the row number.

```{r Q6}
p1 <- ggplot(data = cardiac) + 
  geom_point(aes(x = bmi, y = as.numeric(rownames(cardiac)))) +
  ylab("")
p2 <- ggplot(data = cardiac) +
  geom_point(aes(x = systolic, y = as.numeric(rownames(cardiac)))) +
  ylab("") 
p3 <- ggplot(data = cardiac) +
  geom_point(aes(x = tchol, y = as.numeric(rownames(cardiac)))) +
  ylab("")
p4 <- ggplot(data = cardiac) +
  geom_point(aes(x = alcohol, y = as.numeric(rownames(cardiac)))) +
  ylab("")

# arrange all 4 plots in a single graphics device
pall <- grid.arrange(p1, p2, p3, p4, nrow = 2)

# save the plot
ggsave('output/gg_dotplot.pdf', plot = pall, device = 'pdf')
```

\  

7\. The impossible `bmi` value, and what the plot does about it.

```{r Q7}
which(cardiac$bmi > 100)
# [1] 161
cardiac$bmi[161]
# [1] 514.6
cardiac$bmi[cardiac$bmi > 100] <- NA

ggplot(data = cardiac) +
  geom_point(aes(x = bmi, y = as.numeric(rownames(cardiac)))) +
  ylab("")

# As with the base R version, the remaining extreme values - a bmi of 51.46, a
# systolic of 230, alcohol intakes of 64 and 82 units - are all perfectly
# possible and must be left exactly as they are.
```

\  

8\. Histograms, and the effect of changing the bin width. `ggplot2` will warn you that it has picked a bin number for you, which is a good deal more honest than `hist()` doing it silently.

```{r Q8}
ggplot(data = cardiac) + geom_histogram(aes(bmi))
ggplot(data = cardiac) + geom_histogram(aes(systolic))
ggplot(data = cardiac) + geom_histogram(aes(tchol))
ggplot(data = cardiac) + geom_histogram(aes(alcohol))

# experimenting with different bin widths on systolic
b1 <- ggplot(data = cardiac) + geom_histogram(aes(systolic), binwidth = 20) + ggtitle("binwidth: 20")
b2 <- ggplot(data = cardiac) + geom_histogram(aes(systolic), binwidth = 10) + ggtitle("binwidth: 10")
b3 <- ggplot(data = cardiac) + geom_histogram(aes(systolic), binwidth = 5) + ggtitle("binwidth: 5")
b4 <- ggplot(data = cardiac) + geom_histogram(aes(systolic), binwidth = 2) + ggtitle("binwidth: 2")
grid.arrange(b1, b2, b3, b4, nrow = 2)
```

\  

9\. Scatterplots, and transforming a skewed variable.

```{r Q9}
ggplot(data = cardiac, aes(x = systolic, y = diastolic)) +
  geom_point() +
  labs(x = "systolic (mmHg)", y = "diastolic (mmHg)")

ggplot(data = cardiac, aes(x = hdlchol, y = triglyceride)) +
  geom_point() +
  labs(x = "HDL cholesterol (mmol/l)", y = "triglyceride (mmol/l)")

cardiac$alcohol_sqrt <- sqrt(cardiac$alcohol)
cardiac$alcohol_log  <- log(cardiac$alcohol)

h1 <- ggplot(data = cardiac) + geom_histogram(aes(alcohol)) + ggtitle("untransformed")
h2 <- ggplot(data = cardiac) + geom_histogram(aes(alcohol_sqrt)) + ggtitle("square root")
h3 <- ggplot(data = cardiac) + geom_histogram(aes(alcohol_log)) + ggtitle("natural log")
grid.arrange(h1, h2, h3, nrow = 1)

# ggplot2 is more explicit than base R about the log problem: it prints
# "Removed 58 rows containing non-finite values" rather than quietly dropping
# them. The warning is the useful part - do not learn to ignore it.
```

\  

10\. Boxplots and violin plots by smoking status. Here `ggplot2` has a real advantage: the violin plot needs no extra package, and you can lay the boxplot on top of it.

```{r Q10}
ggplot(data = cardiac, aes(x = Fsmoking, y = tchol)) +
  geom_boxplot() +
  labs(x = "smoking status", y = "total cholesterol (mmol/l)")

ggplot(data = cardiac, aes(x = Fsmoking, y = tchol)) +
  geom_violin(fill = "lightblue") +
  geom_boxplot(width = 0.1) +
  labs(x = "smoking status", y = "total cholesterol (mmol/l)")
```

\  

11\. A pairs plot. `GGally::ggpairs()` is the `ggplot2` equivalent of `pairs()`, and gives you the histograms and correlations without having to write panel functions.

```{r Q11}
ggpairs(cardiac[, c("age", "systolic", "diastolic", "tchol", "hdlchol",
                    "triglyceride", "bmi")])

# coloured by a factor, which is where ggpairs earns its keep
ggpairs(cardiac[, c("systolic", "diastolic", "tchol", "bmi", "Fsex")],
        aes(colour = Fsex, alpha = 0.5))
```

\  

The `ggplot2` version of the Exercise 6 communication plot is at the end of [Exercise 6](exercise_6.html){target="_blank"} itself.

\  

End of Exercise 5 (optional)
