---
title: 'Exercises'
output: 
  html_document:
    toc: false
    code_folding: hide
---


```{r setup, echo=FALSE, purl = FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, eval = FALSE, cache = FALSE,
                      results = 'hold', fig.show = 'hold')
options(width = 80)
SOLUTIONS <- FALSE
```

```{r sandbox, echo=FALSE, eval=TRUE, purl=FALSE}
# Not part of the exercise. When this file is knitted as the solutions the chunks
# below are actually run, so the plots appear on the page. They need a
# 'data/cardiacdata.txt' and an 'output' directory, which students create for
# themselves and which deliberately do not exist in this repository. Build both in
# a temporary sandbox and point knitr at it, so the paths shown in the solutions
# are the paths students really type. _data_raw/cardiacdata_published.txt is the
# tab delimited twin of data/cardiacdata.xlsx; regenerate it if the spreadsheet
# ever changes.
if (SOLUTIONS) {
  .sandbox <- file.path(tempdir(), "ex5_sandbox")
  dir.create(file.path(.sandbox, "data"), recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(.sandbox, "output"), showWarnings = FALSE)
  file.copy("_data_raw/cardiacdata_published.txt",
            file.path(.sandbox, "data", "cardiacdata.txt"), overwrite = TRUE)
  knitr::opts_knit$set(root.dir = .sandbox)
}
```

\  

## Exercise 5: Visualising data using base R

\  

Read [Chapter 4](https://intro2r.com/graphics_r.html) to help you complete the questions in this exercise.

\  

1\. As in previous exercises, either create a new R script or continue with your previous R script in your RStudio Project. Again, make sure you include any metadata you feel is appropriate (title, description of task, date of creation etc) and  don't forget to comment out your metadata with a `#` at the beginning of the line. 

\  

2\. You do not need to download anything for this exercise. You already have *'cardiacdata.txt'* in the `data` directory of your RStudio Project, from Exercise 3. If for some reason you haven't, go back to the **[<i class="fa fa-download"></i> Data](data.html)** link, download *'cardiacdata.xlsx'* and save it as a tab delimited file called *'cardiacdata.txt'* as you did before.

\  

3\. A quick reminder of what these data are. They come from a cohort study of the risk factors for cardiovascular disease, in which 163 adults aged between 55 and 75 were examined and a range of measurements taken: age and sex, systolic and diastolic blood pressure (mmHg), body mass index (kg m^-2^), total cholesterol, HDL cholesterol and triglyceride (all mmol l^-1^), units of alcohol drunk in the previous week, and smoking status. Note that you are importing the original file again, not the cleaned version you exported at the end of Exercise 4. Starting from the raw data every time, and doing your cleaning in a script, is what makes an analysis reproducible.

\  

4\. Import the *'cardiacdata.txt'* file into R using the `read.table()` function and assign it to a variable named `cardiac`. Use the `str()` function to display the structure of the dataset. As you found in Exercise 3, the `sex` and `smoking` variables are coded as integers, but they are really categories. Create a new variable in the `cardiac` dataframe for each of them, recoded as a factor with meaningful labels (`sex` is 1 for Female and 2 for Male; `smoking` is 1 for Current, 2 for Ex and 3 for Never). Use the `str()` function again to check the coding of these new variables. Almost every plotting function in this exercise treats a factor differently from a number, so this step is not optional housekeeping - get it wrong and your boxplots will come out as nonsense.

```{r Q4, echo=SOLUTIONS, eval=SOLUTIONS}
cardiac <- read.table('data/cardiacdata.txt', header = TRUE, sep = "\t", stringsAsFactors = TRUE)

str(cardiac)

# recode the two categorical variables as factors, keeping the originals
cardiac$Fsex <- factor(cardiac$sex, levels = c(1, 2),
                       labels = c("Female", "Male"))

cardiac$Fsmoking <- factor(cardiac$smoking, levels = c(1, 2, 3),
                           labels = c("Current", "Ex", "Never"))

str(cardiac)

#  $ Fsex    : Factor w/ 2 levels "Female","Male": 2 2 2 2 2 2 2 1 1 2 ...
#  $ Fsmoking: Factor w/ 3 levels "Current","Ex",..: NA NA NA NA NA NA NA 1 1 1 ...
```

\  

5\. How many patients are there in each combination of smoking status and sex (hint: remember the `table()` function?)? Don't forget to use the factor recoded versions of these variables. Is the split between the smoking categories similar for women and men? Are any of the combinations small enough to worry about if you wanted to compare them?

```{r Q5, echo=SOLUTIONS, eval=SOLUTIONS}
table(cardiac$Fsmoking, cardiac$Fsex)

  #           Female Male
  # Current       26   24
  # Ex            15   37
  # Never         37   17

# the pattern is almost reversed between the sexes: most of the men are
# ex-smokers, most of the women have never smoked. The smallest cell has 15
# patients, which is small but not unusable.

# and don't forget the 7 patients with no smoking status recorded
table(cardiac$Fsmoking, cardiac$Fsex, useNA = "ifany")
```

\  

6\. The humble cleveland dotplot is a great way of identifying if you have potential outliers in continuous variables (see [Section 4.2.4](https://intro2r.com/simple-base-r-plots.html#dotcharts)). Create dotplots (using the `dotchart()` function) for the following variables; `bmi`, `systolic`, `tchol` and `alcohol`. Do these variables contain any unusually large or small observations? Don't forget, if you prefer to create a single figure with all 4 plots you can always split your plotting device into 2 rows and 2 columns (see [Section 4.4](https://intro2r.com/mult_graphs.html#mult_graphs) of the book).

```{r Q6, echo=SOLUTIONS, eval=SOLUTIONS, fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
dotchart(cardiac$bmi, main = "bmi")
dotchart(cardiac$systolic, main = "systolic")
dotchart(cardiac$tchol, main = "total cholesterol")
dotchart(cardiac$alcohol, main = "alcohol")

# the bmi plot is the striking one: a single point sits so far to the right
# that every other patient is squashed into a narrow strip on the left. You
# cannot see the shape of the bmi distribution at all, because one value is
# setting the scale for all 163.
```

\  

7\. You met that extreme `bmi` value in Exercise 4, when you checked the ranges with `summary()` and found a maximum of 514.60. A body mass index of 514 is not possible: it is 51.46 with the decimal point in the wrong place. What the dotplot adds is not the discovery, but the damage. One bad value has made the plot useless for looking at the other 162 patients. Use the `which()` function to identify which observation it is, confirm the value with the square bracket `[ ]` notation, set it to `NA` as you did in Exercise 4, and then redraw the `bmi` dotchart. Now look again at all four plots from Q6. Several points still sit well away from the rest. Which ones, and what should you do about them?

```{r Q7, echo=SOLUTIONS, eval=SOLUTIONS}
which(cardiac$bmi > 100)
cardiac$bmi[161]
cardiac$bmi[cardiac$bmi > 100] <- NA

dotchart(cardiac$bmi, main = "bmi")

# Now the bmi plot is readable, and you can see the distribution properly:
# most patients between about 20 and 30, thinning out to a handful above 35,
# with one patient at 51.46.

# What else stands out? One patient with a bmi of 51.46, one with a systolic
# pressure of 230 mmHg, and in the alcohol plot a handful of patients reporting
# 41, 64 and 82 units in a week when the median is 2.

# What should you do about them? NOTHING. Every one of those values is
# perfectly possible. A BMI of 51 is severe obesity, a systolic pressure of 230
# is a hypertensive crisis, and 82 units a week is a great deal of alcohol but
# people really do drink that much. These are not errors, they are patients.
# The difference between this question and the last one is the difference
# between a value that CANNOT be right and a value you did not expect, and only
# the first of those is yours to change. Deleting inconvenient data because it
# looks untidy is scientific fraud.
```

\  
 
8\. Histograms are the standard way of looking at the distribution of a continuous variable (see [Section 4.2.2](https://intro2r.com/simple-base-r-plots.html#histograms)). Create histograms for `bmi`, `systolic`, `tchol` and `alcohol`, either separately or in one figure as you did in Q6.

One thing to be careful of is that a histogram can look quite different depending on the number of 'breaks' used. The `hist()` function chooses breaks for you, but they are only a suggestion. Redraw the histogram of `systolic` twice, once with wide bins and once with narrow ones, and compare them. Which one would you show somebody, and what does the difference tell you about how much you should trust the shape of a histogram?

```{r Q8, echo=SOLUTIONS, eval=SOLUTIONS, fig.width=8, fig.height=7}
par(mfrow = c(2, 2))
hist(cardiac$bmi, main = "", xlab = "bmi")
hist(cardiac$systolic, main = "", xlab = "systolic")
hist(cardiac$tchol, main = "", xlab = "total cholesterol")
hist(cardiac$alcohol, main = "", xlab = "alcohol (units/week)")

# wide bins and narrow bins, drawn from exactly the same 163 numbers
par(mfrow = c(2, 1))
hist(cardiac$systolic, xlab = "systolic", main = "bins of 20 mmHg",
     breaks = seq(from = 100, to = 240, by = 20))
hist(cardiac$systolic, xlab = "systolic", main = "bins of 2 mmHg",
     breaks = seq(from = 100, to = 240, by = 2))

# with 20 mmHg bins the distribution looks smooth and roughly symmetric; with
# 2 mmHg bins it looks spiky and full of gaps, because blood pressure is
# recorded in whole even numbers. The data have not changed, only the picture.
# Treat the shape of a histogram as a rough guide, not as evidence.
```

\  

9\. Scatterplots are great for visualising relationships between two continuous variables ([Section 4.2.1](https://intro2r.com/simple-base-r-plots.html#scatterplot)). Before you draw one, stop and think about which variable belongs on which axis. The convention is that the explanatory variable goes on the x axis and the response goes on the y axis, so the choice is not cosmetic. Put them the wrong way round and you have quietly told your reader something you did not mean.

Start with `systolic` and `diastolic` blood pressure. Which would you put where, and why is that a harder question than it first looks? Now plot HDL cholesterol and triglyceride. Which of those two is the more sensible candidate for the x axis, and does the relationship you get make clinical sense to you?

Finally, look at `alcohol`. It is badly skewed, and skew is a nuisance in a plot because most of the points end up squashed into a corner. In Exercise 4 Q5 you took logs of `triglyceride` and it worked nicely. Try two transformations here, a square root with `sqrt()` and a natural log with `log()`, storing each as a new variable in the dataframe, then plot histograms of `alcohol` alongside both. One of the two goes badly wrong. Which one, and why? (hint: what is the smallest value of `alcohol`?)

```{r Q9, echo=SOLUTIONS, eval=SOLUTIONS, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))

# systolic and diastolic: a clear positive relationship, which is no surprise
# because the two numbers are the same measurement taken at two points in the
# heartbeat. Neither one explains the other, so there is no response variable
# here and the choice of axes really is arbitrary. The honest description is
# 'these two go together', not 'this one drives that one'.
plot(cardiac$systolic, cardiac$diastolic,
     xlab = "systolic (mmHg)", ylab = "diastolic (mmHg)")

# triglyceride and HDL cholesterol: negative, and a well known pattern. If you
# are going to treat one as explanatory it should be triglyceride, because a
# raised triglyceride is generally understood to drive HDL down rather than the
# other way round, so triglyceride belongs on the x axis. Do not lean on that
# too hard though. This is a cross sectional study, both are markers of the same
# underlying metabolic state, and no scatterplot can tell you which came first.
plot(cardiac$triglyceride, cardiac$hdlchol,
     xlab = "triglyceride (mmol/l)", ylab = "HDL cholesterol (mmol/l)")

# transforming a skewed variable
cardiac$alcohol_sqrt <- sqrt(cardiac$alcohol)
cardiac$alcohol_log  <- log(cardiac$alcohol)

par(mfrow = c(1, 3))
hist(cardiac$alcohol, main = "untransformed", xlab = "alcohol")
hist(cardiac$alcohol_sqrt, main = "square root", xlab = "sqrt(alcohol)")
hist(cardiac$alcohol_log, main = "natural log", xlab = "log(alcohol)")

sum(cardiac$alcohol == 0)   # 58

# The log fails. 58 of the 163 patients reported drinking no alcohol at all, and
# log(0) is -Inf, so R quietly drops more than a third of your data from the
# plot with nothing more than a warning. The square root works, because sqrt(0)
# is 0, and it pulls the long tail in nicely.

# Triglyceride logged cleanly in Exercise 4 because it has no zeros in it.
# Alcohol does. Check the minimum before you take the log of anything.
```

\  

10\. When visualising differences in a continuous variable between levels of a factor (categorical variable) then a boxplot is your friend (avoid using bar plots - Google 'bar plots are evil' for more info). Create a boxplot to visualise the differences in total cholesterol at each level of smoking status (don't forget to use the recoded version of this variable you created in Q4). Include x and y axis labels in your plot. Make sure you understand the anatomy of a boxplot before moving on - please ask if you're not sure (also see [Section 4.2.3](https://intro2r.com/simple-base-r-plots.html#box-and-violin-plots) of the book).

An alternative to the boxplot is the violin plot, which combines a boxplot with a [kernel density plot](https://www.statmethods.net/graphs/density.html) and shows you the shape of the distribution rather than just its quartiles. You will first need to install the `vioplot` package from CRAN and make it available with `library(vioplot)`. The `vioplot()` function then works in much the same way as `boxplot()`. Draw the same comparison as a violin plot and compare the two.

```{r Q10, echo=SOLUTIONS, eval=SOLUTIONS, tidy = TRUE}
# note: Fsmoking is the recoded smoking variable created in Q4
boxplot(tchol ~ Fsmoking, data = cardiac, xlab = "smoking status", ylab = "total cholesterol (mmol/l)")

# violin plot. install.packages("vioplot") first if you have not already
library(vioplot)
vioplot(tchol ~ Fsmoking, data = cardiac, xlab = "smoking status", ylab = "total cholesterol (mmol/l)", col = "lightblue")

# the three groups look remarkably similar. That is a perfectly good result:
# a plot that shows you there is nothing much to see has still told you
# something, and is a lot more honest than hunting for a variable that does
# show a difference.
```

\  

11\. To explore the relationships between several continuous variables at once it's hard to beat a pairs plot. Create a pairs plot for the variables; `age`, `systolic`, `diastolic`, `tchol`, `hdlchol`, `triglyceride` and `bmi` (see [Section 4.2.5](https://intro2r.com/simple-base-r-plots.html#pairs-plots) of the book for more details). If it looks a little cramped in RStudio then click on the 'zoom' button in the plot viewer to see a larger version.

One of the useful things about `pairs()` is that you can choose what goes into each panel. Modify your plot to add a smoother (a wiggly line) to the lower triangle, which makes the shape of each relationship much easier to see. The book and `?pairs` show you how. Then, just by looking, which pair of variables is most strongly related, and which variable is related to almost nothing?

```{r Q11, echo=SOLUTIONS, eval=SOLUTIONS, fig.width=9, fig.height=9, tidy = TRUE}
plot_vars <- c("age", "systolic", "diastolic", "tchol", "hdlchol", "triglyceride", "bmi")

# vanilla pairs plot
pairs(cardiac[, plot_vars])

# panel.smooth adds the wiggly line. Note that it keeps its dot: it is a base R
# function, not something you wrote
pairs(cardiac[, plot_vars], lower.panel = panel.smooth)

# systolic and diastolic are the most strongly related pair, which is no
# surprise, and hdlchol against triglyceride is the next most obvious, sloping
# the other way. age is related to almost nothing here, and that is worth
# thinking about: everyone in this study is between 55 and 75, so there simply
# is not enough spread in age for a relationship to show itself. A variable can
# look unimportant purely because of who was recruited.
```

\  

11\. (optional) `pairs()` will put whatever you like in the upper triangle, including the correlation coefficients themselves, but there is no ready made function for it so you have to write one. If you fancy a go, `?pairs` has an example you can adapt. Do the numbers agree with what you concluded by eye?

```{r Q11b, echo=SOLUTIONS, eval=SOLUTIONS, fig.width=9, fig.height=9, tidy = TRUE}
panel_cor <- function(x, y, digits = 2, ...) {
  usr <- par("usr"); on.exit(par(usr = usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use = "pairwise.complete.obs")
  text(0.5, 0.5, format(r, digits = digits), cex = 1.4)
}

pairs(cardiac[, plot_vars], upper.panel = panel_cor, lower.panel = panel.smooth)

# systolic with diastolic is the strongest at r = 0.56, hdlchol with
# triglyceride the next at r = -0.43, and age never gets above 0.21 with
# anything. The eye had it right.
```

\  

12\. Every plot so far has appeared in the RStudio plot pane, which is fine while you are exploring but no use at all when you need a figure in a Word document, a PowerPoint slide or a report. Getting a plot out of R is a three step sandwich. You open a graphics device that is a file rather than a window, you draw the plot exactly as you have been doing, and then you close the device with `dev.off()`. Nothing appears on screen while you do it, and if you forget the `dev.off()` your file will be unreadable (see [Section 4.5](https://intro2r.com/export_plots.html#export_plots) of the book).

Redraw your boxplot from Q10 and save it twice into the `output` directory you created in Exercise 1, once as a pdf using `pdf()` and once as a png using `png()`. Open both files outside RStudio and try dropping the png into a Word document. Which format would you use for what, and why? (hint: enlarge each of them a long way and see what happens)

```{r Q12, echo=SOLUTIONS, eval=SOLUTIONS, results='hide'}
# pdf is a vector format: the file stores the instructions for drawing the plot,
# so it stays perfectly sharp however far you enlarge it. Sizes are in inches.
pdf('output/ex5_cholesterol.pdf', width = 7, height = 5)
boxplot(tchol ~ Fsmoking, data = cardiac,
        xlab = "smoking status", ylab = "total cholesterol (mmol/l)")
dev.off()

# png is a bitmap: a grid of pixels, which goes blurry when enlarged. Word and
# PowerPoint handle it happily though. Always set res, because the default of
# 72 dpi looks fine on screen and disappointing on a slide or on paper.
png('output/ex5_cholesterol.png', width = 7, height = 5, units = "in", res = 300)
boxplot(tchol ~ Fsmoking, data = cardiac,
        xlab = "smoking status", ylab = "total cholesterol (mmol/l)")
dev.off()

# check they are really there. Your output directory should now contain
# ex5_cholesterol.pdf and ex5_cholesterol.png
list.files('output')

# tiff() works in exactly the same way and is what journals usually ask for.

# Rule of thumb: pdf if it is going to be printed or enlarged, png at 300 dpi
# if it is going into Word or PowerPoint, tiff if a journal asks for it.
# Whichever you choose, save it from R. Never screenshot the plot pane, and
# never paste a plot in at 72 dpi and hope nobody notices.
```

\  

End of Exercise 5
