---
title: 'Exercises'
output: 
  html_document:
    toc: false
    code_folding: hide
---


```{r setup, echo=FALSE, purl = FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, eval = FALSE, cache = FALSE)
options(width = 80)
SOLUTIONS <- TRUE
```

\  

## Exercise 6: Graphics for communication

\  

Read [Chapter 5](https://intro2r.com/graphics_r.html) to help you complete the questions in this exercise.

\  

**Before the session.** This exercise uses the `ggplot2` package, which doesn't come with R, so please install it before you arrive. Run `install.packages("ggplot2")` in the console. You only ever need to do that once on a given computer, but you will need a `library(ggplot2)` line at the top of any script that uses it.

\  

Everything you plotted in Exercise 5 was for **you**. Exploratory plots are quick, ugly and disposable, and their only job is to show you what's in your data. It doesn't matter that the axis label says `cardiac$tchol` because you're the only person who'll ever see it. This exercise is about the other kind of plot, the one that goes on a poster or in a report, where the reader isn't you, has about fifteen seconds, and won't ask you what the axis means. Nearly all of the work is in the details you skipped last week.

We'll also change graphics package, and it's worth saying why. Base R, which you used in Exercise 5, is ideal for exploring. It's already installed, a plot is one line long, and you throw it away thirty seconds later. For a figure somebody else has to read, `ggplot2` is the better tool. It builds a plot up in layers, it works out the legend for you, and it lets you say what you want rather than how to draw it, which leaves you free to think about the decisions this exercise is really about. A reasonable rule of thumb is base R while you're looking and ggplot2 when you're showing.

You will also switch datasets. From here on the course uses data from the [Scottish Public Health Observatory](https://www.scotpho.org.uk/), which is the same source you will use for your own assessment.

\  

1\. As in previous exercises, either create a new R script or continue with your previous R script in your RStudio Project. Again, make sure you include any metadata you feel is appropriate (title, description of task, date of creation etc) and  don't forget to comment out your metadata with a `#` at the beginning of the line. 

\  

2\. Download the data file *'scotpho_alcohol_admissions.txt'* from the **[<i class="fa fa-download"></i> Data](data.html)** link and save it to your `data` directory. This one is already a tab delimited text file, so there is no conversion to do. Import it with `read.table()`, remembering to write out your `header`, `sep` and `stringsAsFactors` arguments, and assign it to a variable called `scotpho`. Then make `ggplot2` available for the rest of the session with `library(ggplot2)`.

```{r Q2, echo=SOLUTIONS}
library(ggplot2)

scotpho <- read.table('data/scotpho_alcohol_admissions.txt', header = TRUE, sep = "\t", stringsAsFactors = TRUE)
```

\  

3\. These data are an extract from the ScotPHO online profiles tool. The indicator is *alcohol-related hospital admissions*, and the `measure` column is an age-sex standardised rate per 100,000 population: in other words, how many people per 100,000 were admitted to hospital for a reason related to alcohol, adjusted so that areas with different age and sex profiles can be compared fairly. The extract covers each of the 32 Scottish council areas plus Scotland as a whole (`area_name`, `area_code`, `area_type`), for each year from 2010 to 2019 (`year`). Because the rate is an estimate rather than a count, it comes with a confidence interval (`lower_confidence_interval`, `upper_confidence_interval`), which gives you a sense of how precise the estimate is.

\  

4\. Before plotting anything, get your bearings, exactly as you did in Exercise 3. Use `str()` and `summary()` on `scotpho`. How many rows are there? Use `table()` on `area_type` to see how the rows split between council areas and Scotland. Then check something that matters a great deal before you draw a line through points: does every area have a value for every year? (hint: `table()` on `area_name` will tell you how many rows each area has). What would your plot look like if one area were missing 2015?

```{r Q4, echo=SOLUTIONS}
str(scotpho)
summary(scotpho)

nrow(scotpho)            # 330

table(scotpho$area_type)
# Council area     Scotland 
#          320           10 

# 33 areas, 10 years each. Every area is complete:
table(scotpho$area_name)
unique(table(scotpho$area_name))   # 10 - so every area has all 10 years

range(scotpho$year)      # 2010 2019

# If an area were missing a year, the line would simply be drawn straight
# through the gap, joining the points either side of it. Nothing would warn
# you, and the plot would imply data you do not have.
```

\  

5\. Right, let's draw something. Start with the simplest possible communication plot: how have alcohol-related admissions across Scotland as a whole changed over the ten years? Extract just the rows where `area_name` is "Scotland", using the square bracket notation from Exercise 3, and put them in a dataframe called `scot`.

Now for your first ggplot. Every ggplot is built out of the same three pieces (Section [5.2.1](https://intro2r.com/the-start-of-the-end.html#the-purest-of-ggplots) of the book walks through them slowly):

- the **data**, the dataframe you're plotting
- the **aesthetics**, written inside `aes()`, which say which column of that dataframe goes on which part of the plot. Here it's `year` along the x axis and `measure` up the y axis
- a **geom**, which is the thing actually drawn. `geom_line()` draws a line and `geom_point()` draws points, and you can have both

You join them together with `+`, one layer per line. Here is the shape of it, with the parts you need to change in capitals:

```
ggplot(data = YOUR_DATAFRAME, aes(x = YOUR_X_COLUMN, y = YOUR_Y_COLUMN)) +
  geom_line() +
  geom_point()
```

Fill it in and run it. Then look hard at what ggplot has given you by default, and list everything a stranger would not understand about it.

```{r Q5, echo=SOLUTIONS}
scot <- scotpho[scotpho$area_name == "Scotland", ]

ggplot(data = scot, aes(x = year, y = measure)) +
  geom_line() +
  geom_point()

# What is wrong with this plot for anyone other than you?
#  - the axes are labelled year and measure, which are column names rather
#    than English
#  - there are no units anywhere, so 673 could be anything
#  - there is no title, so the reader does not know what is being counted
#  - the x axis is broken at 2010, 2012.5, 2015 and 2017.5, and there is no
#    such year as 2012.5
#  - the y axis runs from about 665 to 764 rather than from zero, which
#    exaggerates the decline
#  - there is nothing to say where the data came from
#
# The plot is not wrong. It is just useless to anybody who is not already
# holding the dataset.
```

\  

6\. A single line rarely tells the whole story. Scotland's average hides enormous variation between council areas, so let's add two of them, "Glasgow City", which has the highest rate in Scotland, and "Aberdeenshire", which has one of the lowest.

This is where ggplot starts to earn its keep. In base R you would pull each area into its own dataframe and draw three separate lines, keeping the colours and the legend in step by hand. ggplot works the other way round. You give it one dataframe holding all three areas, tell it which column says which area a row belongs to, and it draws one line per area and builds the legend for you.

a) First get those three areas into a single dataframe. You could do it with three conditions joined by `|`, the 'or' operator, but there's a tidier way. The `%in%` operator takes a value on its left and a vector on its right and asks whether that value appears anywhere in the vector. Hand it a whole column and it does that for every row in turn, giving you back a `TRUE` or a `FALSE` for each one, which is exactly what the square brackets want. It's the same job you did with `==` and `&` in Exercise 3, Q10, just against a list of possibilities rather than a single one.

So `scotpho$area_name %in% areas` gives you 330 `TRUE`s and `FALSE`s, one per row of `scotpho`. Make a vector called `areas` holding the three area names, use it to extract those rows into a dataframe called `three_areas`, and check with `nrow()` that you have the 30 rows you expect. The nice thing about writing it this way is that it doesn't get any longer if you decide you want six areas instead of three.

b) Now plot it, with the same three pieces as Q5 and one addition. Inside `aes()`, alongside your `x` and `y`, add `colour = area_name`. That single argument is what tells ggplot the rows belong to three different areas, and it does the rest, a line each and a legend down the side.

c) Look at the legend. The areas are in alphabetical order, because that's the order R put the factor levels in when it read the file, and it's rarely the order you'd choose for a reader. You can set the order yourself by rebuilding the factor with the levels in the order you want:

```
three_areas$area_name <- factor(three_areas$area_name, levels = areas)
```

Redraw the plot and the legend follows. What do the three lines tell you that the Scotland line on its own did not?

```{r Q6, echo=SOLUTIONS}
# a) the three areas in one dataframe
areas <- c("Scotland", "Glasgow City", "Aberdeenshire")

three_areas <- scotpho[scotpho$area_name %in% areas, ]

nrow(three_areas)   # 30 - three areas, ten years each

# b) one geom_line(), three lines, and a legend you did not have to write
ggplot(data = three_areas, aes(x = year, y = measure, colour = area_name)) +
  geom_line(linewidth = 1)

# c) the legend comes out alphabetically, Aberdeenshire, Glasgow City,
# Scotland. Rebuilding the factor with the levels in the order you want puts
# the national figure first and the two council areas either side of it.
three_areas$area_name <- factor(three_areas$area_name, levels = areas)

ggplot(data = three_areas, aes(x = year, y = measure, colour = area_name)) +
  geom_line(linewidth = 1)

# Glasgow City is roughly twice the Scottish average and Aberdeenshire is
# roughly half of it. In 2019 the rates were 1169, 673 and 314 admissions per
# 100,000. The national line, on its own, describes almost nobody.
```

\  

7\. Now think about those colours. ggplot's defaults are perfectly pleasant, but they aren't chosen with anybody's eyesight in mind. Roughly 1 in 12 men has some form of colour vision deficiency, most commonly difficulty distinguishing red from green, and red and blue are not a safe pairing either. Posters also get photocopied, printed in black and white, and viewed on projectors that wash the colour out.

a) Redraw your plot with a colour-blind friendly palette. Base R has one built in, and `palette.colors(3, palette = "Okabe-Ito")` gives you three colours from it. Assign those to `cols`, then hand them to ggplot by adding a `scale_colour_manual(values = cols)` layer on the end of your plot. A scale layer is how you overrule any of the decisions ggplot has made for you (Section [5.2.4](https://intro2r.com/the-start-of-the-end.html#make-it-your-own)).

b) Now test it. Draw it a third time, replacing those colours with their greyscale equivalents, which are `grey(c(0, 0.64, 0.62))`, and see whether the three lines are still tellable apart. If they aren't, what could you change other than the colour?

c) Fix it. Colour on its own is never enough, so give the reader a second, redundant cue. Add `linetype = area_name` inside `aes()`, next to the `colour = area_name` you already have, and each area gets its own style of line as well as its own colour. Notice that ggplot works out that the two are describing the same thing and gives you one legend rather than two.

```{r Q7, echo=SOLUTIONS}
# a) a colour-blind friendly palette
cols <- palette.colors(3, palette = "Okabe-Ito")
cols
# "#000000" "#E69F00" "#56B4E9"   black, orange, sky blue

ggplot(data = three_areas, aes(x = year, y = measure, colour = area_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols)

# b) the same plot in greyscale
greys <- grey(c(0, 0.64, 0.62))

ggplot(data = three_areas, aes(x = year, y = measure, colour = area_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = greys)

# The orange and the sky blue are different enough on screen, but in grey they
# come out at 0.64 and 0.62, which is practically the same. Anyone printing
# your poster in black and white cannot tell Glasgow from Aberdeenshire.

# c) a second cue, so the plot still works with no colour at all
ggplot(data = three_areas,
       aes(x = year, y = measure, colour = area_name, linetype = area_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols)

# mapping both colour and linetype to the same variable gives you a single
# legend showing both, which is what you want. Try it with greys instead of
# cols and you will find the plot still reads perfectly well.
```

\  

8\. Time to fix the words. Every plot that leaves your computer needs axis labels a non-specialist can read, including units; a title that says what the reader is looking at; and a note saying where the data came from. Write the labels for someone who has never heard the phrase 'age-sex standardised'. Your reader is a public health colleague or a member of the public, not a statistician. Keep this version of the code safe, because you'll need it again in Q9.

In ggplot all of the words live in one `labs()` layer, which takes `x`, `y`, `title` and `caption` (Section [5.2.6](https://intro2r.com/the-start-of-the-end.html#prettification)). The caption is where the source note goes. While you're in there, `colour = NULL` and `linetype = NULL` inside the same `labs()` will get rid of the 'area_name' heading above the legend, which is another column name your reader doesn't need to see.

Three more things to sort out, each of them a layer.

a) The x axis is still labelled with half years. Add `scale_x_continuous(breaks = seq(2010, 2019, by = 2))` to put the breaks on real years.

b) Think about the y axis. Should it start at zero? There is no single right answer, but you should be able to justify whichever you choose. If you decide it should, `scale_y_continuous(limits = c(0, 1600))` will do it (Section [5.3.2](https://intro2r.com/tips.html#axis-limits-and-zooms)).

c) Finally, the look of the thing. ggplot's grey panel is fine on screen and muddy in print. A **theme** controls everything about the plot that isn't your data, and there are several ready-made ones (Section [5.2.5](https://intro2r.com/the-start-of-the-end.html#setting-the-theme)). Add `theme_minimal()` and see what you think. If you'd rather the legend sat under the plot than beside it, `theme(legend.position = "bottom")` will move it, and adding it after `theme_minimal()` means it isn't overwritten.

```{r Q8, echo=SOLUTIONS}
ggplot(data = three_areas,
       aes(x = year, y = measure, colour = area_name, linetype = area_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 2)) +
  scale_y_continuous(limits = c(0, 1600)) +
  labs(x = "Year",
       y = "Hospital admissions per 100,000 people",
       title = "Alcohol-related hospital admissions, 2010 to 2019",
       caption = "Source: Scottish Public Health Observatory. Rates are age-sex standardised.",
       colour = NULL, linetype = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", plot.caption = element_text(hjust = 0))

# plot.caption = element_text(hjust = 0) pushes the source note over to the
# left. ggplot right aligns it by default, which looks like an afterthought.

# Should the y axis start at zero?
# Starting at zero, as here, shows the true relative size of the difference
# between areas: Glasgow really is about four times Aberdeenshire. Starting at
# the minimum instead would fill the panel with the year to year wiggles and
# make a modest national decline look dramatic. For a rate like this, where
# zero is meaningful and the comparison between areas is the point, starting at
# zero is the honest choice. For something like average age, where zero is
# nowhere near the data, it would be absurd. Decide, and be able to say why.
```

\  

9\. Finally, get the plot out of R at a size and resolution fit for a poster. Screen defaults are no use here, and a plot that looks fine in the RStudio plot pane will be a blurry mess blown up to A1.

This is easier if you give the plot a name first. Assign the whole thing from Q8 to an object called `admissions_plot`, exactly as you'd assign anything else in R. Nothing will appear when you run it, because you've stored the plot rather than drawn it; type `admissions_plot` on a line of its own to see it.

Now save it with `ggsave()`, which is ggplot's own version of the `pdf()` and `dev.off()` sandwich you met in Exercise 5, Q12. There's no device to close, because `ggsave()` opens and closes it for you, and it works out which format you want from the file extension you give it (Section [5.2.6](https://intro2r.com/the-start-of-the-end.html#prettification)).

Save your plot twice into the `output` directory you created in Exercise 1, once as `ex6_admissions.pdf` and once as `ex6_admissions.png`. Use `width = 10` and `height = 5.625`, which is a 16:9 shape and what most poster and slide templates expect, and for the png set `dpi = 300`, which is the usual minimum for print. Open both files and check them at full size before you believe them.

**Keep the png.** You'll put it into a report in Exercise 7 and then into a repository in Exercise 9, so it needs to still be there in a fortnight.

```{r Q9, echo=SOLUTIONS}
# give the plot a name. Nothing is drawn until you ask for it by name
admissions_plot <- ggplot(data = three_areas,
       aes(x = year, y = measure, colour = area_name, linetype = area_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 2)) +
  scale_y_continuous(limits = c(0, 1600)) +
  labs(x = "Year",
       y = "Hospital admissions per 100,000 people",
       title = "Alcohol-related hospital admissions, 2010 to 2019",
       caption = "Source: Scottish Public Health Observatory. Rates are age-sex standardised.",
       colour = NULL, linetype = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", plot.caption = element_text(hjust = 0))

admissions_plot     # draw it in the plot pane

# pdf: a vector format, so it stays sharp at any size. Sizes are in inches.
ggsave('output/ex6_admissions.pdf', plot = admissions_plot,
       width = 10, height = 5.625, units = "in")

# png: pixels, so the resolution matters. 10 x 5.625 inches at 300 dpi gives
# you a 3000 x 1687 pixel image that will print cleanly.
ggsave('output/ex6_admissions.png', plot = admissions_plot,
       width = 10, height = 5.625, units = "in", dpi = 300)

# Use the pdf for anything going to a printer and the png for anything going
# into a Word document, PowerPoint or a web page. If your poster template asks
# for a particular figure size, set width and height to that size here rather
# than resizing the image afterwards, which is what makes text look squashed
# or fuzzy.

# If you leave out plot = admissions_plot, ggsave() saves the last plot you
# drew, which is usually what you wanted but not always. Naming it is safer.
```

\  

End of Exercise 6
