---
title: 'Exercises'
output: 
  html_document:
    toc: false
    code_folding: hide
---


```{r setup, echo=FALSE, purl = FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, eval = FALSE, cache = FALSE)
options(width = 80)
SOLUTIONS <- TRUE
```

\  

## Exercise 3: Importing, inspecting and subsetting data

\  

Read [Chapter 3](https://intro2r.com/data_r.html) to help you complete the questions in this exercise.

\  

1\. As in previous exercises, either create a new R script or continue with your previous R script in your RStudio Project. Again, make sure you include any metadata you feel is appropriate (title, description of task, date of creation etc) and  don't forget to comment out your metadata with a `#` at the beginning of the line.

\  

2\. If you haven't already, download the data file *'cardiacdata.xlsx'* from the **[<i class="fa fa-download"></i> Data](data.html)** link and save it to the `data` directory you created in Exercise 1 in your RStudio project. You will use this same dataset for the next three practicals, so it is worth putting it somewhere sensible now.

\  

3\. Notice that the file you have just downloaded is a spreadsheet. A spreadsheet is a program's own format, and R would much rather be given plain text. Open the *'cardiacdata.xlsx'* file in Microsoft Excel (or even better use an open source equivalent - [LibreOffice](https://www.libreoffice.org/download/download-libreoffice/) is a good free alternative) and save it as a tab delimited file type (see [Section 3.3.1](https://intro2r.com/importing-data.html#saving-files-to-import) of the Introduction to R book or watch this [video](https://alexd106.github.io/PU5058/howto.html#import-vid) if you're not sure how to do this). Name the file *'cardiacdata.txt'* and save it to the `data` directory. If you're a Windows user be careful with file extensions (things like *.txt*). By default, Windows doesn't show you the file extension (maybe the boffins at Microsoft don't think you need to know complicated things like this?) so if you enter *'cardiacdata.txt'* as a filename you might end up with a filename *'cardiacdata.txt.txt*'! 

\  

4\. Time for a quick description of the *'cardiacdata.txt'* dataset to get your bearings. These data come from a cohort study of the risk factors for cardiovascular disease. A group of 163 adults, aged between 55 and 75, were examined at the start of the study and a range of measurements was taken on each of them. Each row of the dataset is one patient, identified by a unique patient number (`patno`), and each column is one of those measurements. As well as their age in years (`age`) and sex (`sex`), the study recorded blood pressure as the usual pair of numbers, systolic and diastolic, in mmHg (`systolic`, `diastolic`); body mass index in kg m^-2^ (`bmi`); three fats measured in a blood sample, all in mmol l^-1^, namely total cholesterol, HDL cholesterol and triglyceride (`tchol`, `hdlchol`, `triglyceride`); how many units of alcohol each person drank in the previous week (`alcohol`); and whether they were a current smoker, an ex-smoker or had never smoked (`smoking`). The structure of these data is known as a rectangular dataset (aka 'tidy' data by the cool kids). Each row is an individual observation, each column a separate variable, and the variable names are contained in the first row of the dataset (aka a header). Unlike the tidy examples in textbooks, though, this dataset is not complete. A few values are missing, and wherever a measurement was not recorded the cell in the file holds `NA`, which is how R writes a missing value and what it will expect to find when it imports the file. On top of that, at least one value is present but cannot possibly be correct. Finding both is part of your job over this practical and the next.

\  

5\. Now let's import the *'cardiacdata.txt'* file into R. To do this you will use the workhorse function of data importing,  `read.table()`. This function is incredibly flexible and can import many different file types (take a look at the help file) including our tab delimited file. Rather than relying on the default settings, get into the habit of writing out the arguments you care about every time. `header` tells R whether the first row contains the variable names, `sep` tells it what separates one value from the next, and `stringsAsFactors` tells it whether to convert text to factors. Being explicit costs you a few seconds and saves you a great deal of confusion later. Assign the imported data to a variable with an appropriate name (such as `cardiac`). Take a look at [Section 3.2.2](https://intro2r.com/importing-data.html#import_fnc) of the Introduction to R book or watch this [video](https://alexd106.github.io/PU5058/howto.html#import-vid) if you need any further information.

```{r Q5, echo=SOLUTIONS}
cardiac <- read.table('data/cardiacdata.txt', header = TRUE, sep = "\t", stringsAsFactors = TRUE)
```

\  

6\. Once you've imported your data file nothing much seems to happen (don't worry, this is normal). To examine the contents of the dataframe one option would be to just type the variable name (`cardiac`) into the console. This is probably not a good idea and doesn't really tell you anything about the dataframe other than there's alot of data (try it)! A slightly better option is to use the `head()` function to display the first few rows of your dataframe. Again, this is likely to just fill up your console. A better option would be to use the `names()` function which will return a vector of variable names from your dataset. However, all you get are the names of the variables but no other information. A much, much better option is to use the `str()` function to display the structure of the dataset and a neat summary of your variables. Another advantage is that you can copy this information from the console and paste it into your R script (making sure it's commented) for later reference. 

So, from the output of the `str()` function, how many observations does this dataset have? How many variables? And now the important question: what type of variable does R think `sex` and `smoking` are, and is it right?

```{r Q6, echo=SOLUTIONS}
head(cardiac)         # display the first few rows 
names(cardiac)        # display the variable names
str(cardiac)          # display the structure of the dataframe cardiac

# 'data.frame':	163 obs. of  11 variables:
#  $ patno       : Factor w/ 163 levels "0049B","0052H",..: 6 54 65 69 ...
#  $ age         : num  71 68.2 62.9 69.9 65 ...
#  $ sex         : int  2 2 2 2 2 2 2 1 1 2 ...
#  $ systolic    : int  142 140 156 154 187 123 147 140 131 128 ...
#  $ diastolic   : int  63 78 82 102 131 63 66 67 73 62 ...
#  $ tchol       : num  7.03 5.32 9.34 7.19 8.84 6.17 6.2 6.96 7.02 7.21 ...
#  $ hdlchol     : num  1.4 0.88 0.92 1.31 1.83 1.58 1.65 1.68 1.57 1.08 ...
#  $ triglyceride: num  0.81 3.4 4.67 2.53 1.76 0.73 1.11 0.69 1.29 1.16 ...
#  $ bmi         : num  24.7 26 26.6 26.2 26.1 ...
#  $ alcohol     : int  9 2 7 24 14 8 0 2 0 0 ...
#  $ smoking     : int  NA NA NA NA NA NA NA 1 1 1 ...

# 163 observations and 11 variables.

# R thinks sex and smoking are integers, because that is how they are stored
# in the file: sex is coded 1 and 2, smoking is coded 1, 2 and 3. But they are
# not really numbers, they are categories, and R has no way of knowing that.
# Nothing stops you calculating mean(cardiac$sex) and it returns 1.52 - and that
# number is mostly meaningless. You will fix both of them in the next question.

# patno is a factor with 163 levels, one for every row. A factor with as many
# levels as there are rows is a good sign that you are looking at an identifier
# rather than a variable.
```

\  

7\. Let's do something about the `sex` and `smoking` variables. The way to tell R that a variable holds categories rather than quantities is to make it a **factor**, using the `factor()` function (see `?factor`, and [Section 3.1](https://intro2r.com/data-types.html#data-types) of the Introduction to R book for where factors sit among R's data types). You give `factor()` the codes that appear in the data with the `levels` argument, and what each code means with the `labels` argument. For `sex`, 1 is Female and 2 is Male. For `smoking`, 1 is Current, 2 is Ex and 3 is Never.

Here is the important part. Do **not** overwrite either of them. Create two **new** variables in the `cardiac` dataframe, called for example `Fsex` and `Fsmoking`, and leave the original `sex` and `smoking` variables exactly as they were. Do this, then use `str()` to check that both really are factors with the levels you expect, and that `sex` and `smoking` are still there and still integers.

Why do we create a new variable rather than overwriting the original? Have a think about it before you reveal the answer.

```{r Q7, echo=SOLUTIONS}
# Writing the labels out means you never have to remember the codes again, and
# your output reads properly from here on.
cardiac$Fsex <- factor(cardiac$sex, levels = c(1, 2),
                       labels = c("Female", "Male"))

cardiac$Fsmoking <- factor(cardiac$smoking, levels = c(1, 2, 3),
                           labels = c("Current", "Ex", "Never"))

str(cardiac)

#  $ sex         : int  2 2 2 2 2 2 2 1 1 2 ...
#  $ smoking     : int  NA NA NA NA NA NA NA 1 1 1 ...
#  $ Fsex        : Factor w/ 2 levels "Female","Male": 2 2 2 2 2 2 2 1 1 2 ...
#  $ Fsmoking    : Factor w/ 3 levels "Current","Ex",..: NA NA NA NA NA NA NA 1 1 1 ...

# the 7 patients with no smoking code stay missing in Fsmoking, which is exactly
# right. factor() does not invent a category for them.

# Why new variables rather than overwriting sex and smoking?
#
# 1. The raw codes survive. You can always check your recoding against what was
#    actually in the file, and if you get the labels the wrong way round you can
#    simply redo it. Overwrite, and the codes are gone from your session and the
#    only way back is to import the file again.
#
# 2. The 'F' prefix says at a glance which variable is the factor version. You
#    will use the same convention every time you recode something.

# and note: the originals are still integers, exactly as they were
str(cardiac$sex)
str(cardiac$smoking)
```

\  

8\. You can get another useful summary of your dataframe by using the `summary()` function. This will provide you with some useful summary statistics for each variable. Notice how the type of output depends on whether the variable is a factor or a number: for `Fsex` and `Fsmoking`, the factors you just created, you get a count of patients in each level, whereas for a numeric variable you get the minimum, maximum, mean, median and quartiles. Compare `sex` and `Fsex` in the output and you will see the same information summarised in two very different ways. Another useful feature of the `summary()` function is that it will also count the number of missing values in each variable. Which variables have missing values, and how many? While you are looking at this output, check the maximum value of `bmi`. Does that look like a plausible body mass index to you? Make a note of it and hang on to that thought until Exercise 4.

```{r Q8, echo=SOLUTIONS}
summary(cardiac)

# NOTE: only some of the columns are shown here, and not in their original
# order, to save space

 #     Fsex        tchol           hdlchol       triglyceride        bmi        
 #  Female:78   Min.   : 4.120   Min.   :0.000   Min.   :0.000   Min.   : 17.57  
 #  Male  :85   1st Qu.: 6.140   1st Qu.:1.080   1st Qu.:1.040   1st Qu.: 22.88  
 #              Median : 6.830   Median :1.360   Median :1.380   Median : 25.20  
 #              Mean   : 6.978   Mean   :1.404   Mean   :1.547   Mean   : 28.72  
 #              3rd Qu.: 7.720   3rd Qu.:1.700   3rd Qu.:1.870   3rd Qu.: 28.24  
 #              Max.   :11.660   Max.   :3.000   Max.   :4.670   Max.   :514.60  
 #              NA's   :2        NA's   :2       NA's   :2                       

# Fsex is a factor, so summary() counts patients: 78 women and 85 men. The
# untouched sex column is still an integer, so summary() reports a
# mean of 1.52 for it.

# Four variables have missing values: tchol, hdlchol and triglyceride have 2
# each, and smoking has 7.

# The maximum bmi is 514.60. A body mass index of 514 is not possible; the
# heaviest person ever recorded had a BMI of around 200. Notice that nothing
# went wrong when you imported the file, and R will happily calculate a mean
# from it (28.72, when the median is 25.20). It is simply a wrong number sitting
# quietly in a column of right ones. You will deal with it in Exercise 4.
```

\  

9\. Summarising and manipulating dataframes is a key skill to acquire when learning R. Although there are many ways to do this, we will concentrate on using the square bracket `[ ]` notation which you used previously with vectors. The key thing to remember when using `[ ]` with dataframes is that dataframes have two dimensions (think rows and columns) so you always need to specify which rows and which columns you want inside the `[ ]` (see [Section 3.4.1](https://intro2r.com/wrangling-data-frames.html#positional-indexes) and this [video](https://alexd106.github.io/PU5058/howto.html#dataw-vid) for some additional background information and a few examples). Let's practice. 

a) Extract all observations (remember - rows) from the `cardiac` dataframe and the columns `patno`, `sex`, `smoking` and `tchol` and assign to a variable called `cardiac_risk`. 

b) Next, extract all rows except the first 10 rows and all columns except the last column. Careful here, which column is the last one now? Remember, you can specify the columns you want either by position or by name. Practice both ways. 

```{r Q9, echo=SOLUTIONS}
# all rows and the columns patno, sex, smoking and tchol
cardiac_risk <- cardiac[, c(1, 3, 11, 6)] 
# alternative way of indexing columns with named indexes - much easier to read,
# and it still works if the column order changes
cardiac_risk <- cardiac[, c("patno", "sex", "smoking", "tchol")]    

# excluding first 10 rows and last column using negative indexing. NOTE: the
# last column is now Fsmoking, the second of the two you created in Q7. The
# dataframe has 13 columns rather than the 11 you imported, which is exactly why
# hard coded column numbers are fragile.
cardiac_last <- cardiac[-c(1:10), -13]  
# more general way, and it stays correct however many columns you add
cardiac_last <- cardiac[-c(1:10), -c(ncol(cardiac))] 
# NOTE: negative indexing does NOT work with column names. Uncomment the line
# below and run it to see the error for yourself:
# cardiac_last <- cardiac[-c(1:10), -c("Fsmoking")]
```

\  

10\. In addition to extracting rows and columns from your dataframe by position (as you have just been doing) you can also use conditional statements to select particular rows based on some logical criteria. This is very useful but takes a bit of practice to get used to (see [Section 3.4.2](https://intro2r.com/wrangling-data-frames.html#logical-indexes) for an introduction). Extract rows from your dataframe (all columns by default) based on the following criteria (note: you will need to assign the results of these statements to appropriately named variables, I'll leave it up to you to use informative names!). You now have properly labelled `Fsex` and `Fsmoking` variables, so use those rather than the raw codes. Your conditions will then read like English and you will not have to keep looking up which number means what:  

- systolic blood pressure greater than 160 mmHg
- male patients who have never smoked and whose diastolic pressure is greater than the median diastolic pressure (76 mmHg)
- all patients who are not ex-smokers

```{r Q10, echo=SOLUTIONS, tidy = TRUE}
cardiac_sys160 <- cardiac[cardiac$systolic > 160, ]

cardiac_subset <- cardiac[cardiac$Fsex == "Male" & cardiac$Fsmoking == "Never" & cardiac$diastolic > 76, ]

cardiac_notex <- cardiac[cardiac$Fsmoking != "Ex", ]
```

\  

11\. A really neat feature of conditional statements is that you can put R functions inside them, so the threshold is worked out from the data rather than typed in by hand (see [Section 3.4.2](https://intro2r.com/wrangling-data-frames.html#logical-indexes) again). This is useful becuase if you hard code the median as 76, as you did in Q10, your code becomes silently wrong the moment somebody adds another twenty patients to the study, whereas `median(cardiac$diastolic)` is still right.

Put let's put this into practice by building on your solution to Q10. Write some code to extract all rows from the `cardiac` dataframe with a systolic pressure greater than 160 mmHg and a total cholesterol greater than average (hint: use the `mean()` function inside your conditional statement). Now look carefully at what comes back. Can you see a problem with it? Have a look at [Section 2.3](https://intro2r.com/vectors.html#na_vals) of the Introduction to R book, discuss the cause of this problem with an instructor and explore possible solutions.

```{r Q11, echo=SOLUTIONS, tidy = TRUE}
# This results in a dataframe filled with NAs. Why?
cardiac_new <- cardiac[cardiac$systolic > 160 & cardiac$tchol > mean(cardiac$tchol), ]

# the variable tchol contains 2 NA values. By default the mean function will return an NA.
# use the na.rm argument to ignore NAs
cardiac_new <- cardiac[cardiac$systolic > 160 & cardiac$tchol > mean(cardiac$tchol, na.rm = TRUE), ]  
```

\  

End of Exercise 3
