---
title: 'Setting up Git and GitHub'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

\  

## Before you start

**You do not need this at the start of the course.** It is needed for Exercise 8, and there is no advantage in doing it earlier; the access token you create in Step 5 has a limited life, so setting it up months in advance only means renewing it sooner.

\  

This page gets your computer talking to GitHub. It takes about half an hour and it is
fiddly the first time, which is why we ask you to do it **before** the Exercise 8
practical rather than during it.

Please work through it beforehand. If you hit a real problem, bring it to the start of the
session and we will help you, but we cannot get everybody set up and still have time to
teach anything, so do try first.

You only ever do this once per computer.

\  

## Step 1: Install Git

Git is a separate program from R and RStudio, and you need it installed before RStudio can
use it. There is a fuller version of these instructions in [Appendix B](https://intro2r.com/install-git.html)
of the Introduction to R book.

**Windows.** Download [Git for Windows](https://gitforwindows.org/) and run the installer.
There are a lot of screens and you can accept every default on all of them. The installer
also adds a program called Git Bash, which we will not use on this course but which does no
harm.

**macOS.** The simplest route is to let your Mac install it for you. Open the Terminal
application (press Cmd+Space, type `Terminal`, press Enter), type `git --version` and press
Enter. If Git is not installed, macOS will offer to install the command line developer
tools; accept, and wait for it to finish. This is the only time on this course you will need
the Terminal. If you would rather use an installer, there is one at
[git-scm.com](https://git-scm.com/download/mac).

**Linux.** Git is often installed already. Check with `git --version`. If not, use your
package manager: `sudo apt install git` on Debian or Ubuntu, `sudo dnf install git` on
Fedora.

\  

## Step 2: Create a GitHub account

Git runs on your computer. GitHub is a website that stores copies of your work and is what
makes it a backup rather than just a filing system.

Sign up for a free account at [github.com/join](https://github.com/join). You need a
username, an email address and a strong password. Choose a username you would be happy for
a future employer to see: quite a lot of people put their GitHub profile on a CV.

Two-factor authentication is a good idea and takes about five minutes to set up with an
authenticator app on your phone. It is **not** required for what we are doing, so do not
let it hold you up now. You can add it later from your GitHub account settings.

\  

## Step 3: Tell RStudio where Git is

Open RStudio and go to `Tools` -> `Global Options` -> `Git/SVN`.

Check that **'Enable version control interface for RStudio projects'** is ticked, and that
the **'Git executable'** box has a sensible path in it. On Windows this usually ends in
`git.exe`, on macOS and Linux in `git`.

If the box is empty, click `Browse...` and find the program you installed in Step 1. Restart
RStudio afterwards.

\  

## Step 4: Tell Git who you are

Git labels every change you make with a name and an email address, so you need to give it
those once.

You may have seen this done in a terminal. We will use the `usethis` package instead, so
that everything happens in the R console you already know. Install it first if you have not
already:

```{r step4a, eval=FALSE}
install.packages("usethis")
```

Then:

```{r step4b, eval=FALSE}
library(usethis)

use_git_config(user.name = "Your Name", user.email = "you@example.com")
```

Use the same email address you signed up to GitHub with. The name is what will appear
against your changes, so use your real one.

While you are here, set the default name for the main line of work in new projects:

```{r step4c, eval=FALSE}
use_git_config(init.defaultBranch = "main")
```

GitHub has called this `main` since 2020. Git itself still uses the older name `master`
unless you tell it otherwise, and the mismatch causes confusing errors later. You will meet
both names in the wild.

\  

## Step 5: Create a personal access token

This is the step people get stuck on, so read it carefully. [Section 9.4.5](https://intro2r.com/setup_git.html#github-auth)
of the Introduction to R book covers the same ground.

When you send your work to GitHub, it needs to know it is really you. GitHub **stopped
accepting your account password for this in 2021**. Instead you use a *personal access
token*, which is a long string of characters that works like a password for one computer,
and which you can cancel without changing anything else about your account.

Run:

```{r step5, eval=FALSE}
create_github_token()
```

This opens GitHub in your web browser with a form already filled in with the right
permissions. Before you click the green **Generate token** button at the bottom, change two
things:

1. **Note**: replace whatever is there with something that will still mean something to you
   in six months, such as `PU5058 laptop`.

2. **Expiration**: click the dropdown and choose `Custom`, then pick a date about six months
   from now. GitHub suggests 30 days by default. Thirty days is better security practice,
   but it means your token stops working part way through this course, probably during the
   assessment. We are deliberately trading a little security for not being locked out at the
   worst moment. When you finish this course, go back to shorter-lived tokens.

Now click **Generate token**. GitHub shows you the token exactly once, on a green
background. Copy it now. If you navigate away without copying it, you cannot get it back and
will have to make another one, which is annoying but harmless.

\  

## Step 6: Store the token

Give the token to your computer's password store so you never have to type it again:

```{r step6, eval=FALSE}
gitcreds::gitcreds_set()
```

Paste the token when prompted and press Enter. (You already have `gitcreds` installed:
`usethis` depends on it.)

\  

## Step 7: Check it all worked

```{r step7, eval=FALSE}
git_sitrep()
```

`git_sitrep()` is short for 'Git situation report'. It prints everything about your setup in
one go: which Git you have, the name and email you set in Step 4, and whether it can find
and use your token.

Read the output even though nothing has gone wrong. Knowing what it looks like when things
are healthy is exactly what makes it useful when they are not. The lines to check are your
name and email, and a line confirming a personal access token was found.

If anything looks wrong, go back to the step that set it. If you cannot see what is wrong,
bring the `git_sitrep()` output to the session and we will look at it together.

\  

## When your token expires

Your token will expire eventually. This is normal and nothing is lost when it happens.

The symptom is that pushing or pulling, which worked perfectly well last week, suddenly
fails with a message about authentication. The fix is to repeat Steps 5 and 6:

```{r expired, eval=FALSE}
create_github_token()      # generate a fresh one
gitcreds::gitcreds_set()   # it will offer to replace the old one - say yes
git_sitrep()               # check
```

Your work is on your computer and on GitHub throughout. An expired token is a five minute
annoyance, not a disaster.

\  

## Where to go if you are stuck

The standard reference for R users is [Happy Git and GitHub for the useR](https://happygitwithr.com),
which covers everything on this page in far more detail, including what to do when the
usual routes fail.
