---
title: "Lectures"
output: 
  html_document:
    toc: false
---

\  

Lecture slides that we deliver during our in-person courses to help support you whilst learning R. This material should be used in conjunction with the [Introduction to R book](https://alexd106.github.io/Rbook). 

Either tap on the slides or use your keyboard arrow keys (&#8592; &#8594;) to scroll through the slides. Press 'h' for other slide options.

\  

```{r, reveal, echo=FALSE, eval = isFALSE(show_lecture)}
cat("Lectures will be released as the course progresses")
```

\  

```{r intro-lec, echo=FALSE}
knitr::include_url('https://alexd106.github.io/Rintro_lec/Rintro_lecture#1')
```

\  

```{r data-lec, echo=FALSE}
knitr::include_url('https://alexd106.github.io/Rdata_lec/Rdata_lecture#1')
```

\  

```{r stat-lec, echo=FALSE}
knitr::include_url('https://alexd106.github.io/Rstats_lec/Rstats_lecture#1')
```

\  

When available, follow the links below for PDF versions of old lecture slides.

\  

\  

```{r, echo=FALSE, eval = isTRUE(show_lecture)}
lecture_links <- list(
  list("Introduction to R", "images/r_lecture_thumb.png", "lectures/introduction_to_r_2020.pdf"),
  list("Data in R", "images/data_in_r_thumb.png", "lectures/getting_the_data_in_r.pdf"),
  list("Simple Statistics in R", "images/stats_r_thumb.png", "lectures/statistics_in_r.pdf")
)

for (lecture_link in lecture_links) {
  if (file.exists(lecture_link[[2]]) && file.exists(lecture_link[[3]])) {
    print(thumbnail(lecture_link[[1]], lecture_link[[2]], lecture_link[[3]]))
  }
}
```

\  

\  
