---
title: 'Setting up R markdown'
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

\

## Before you start

You need this for Exercise 7, and you'll need it again for your report assessment, because
both of them produce PDF documents. It takes about fifteen minutes, most of which is a
download running by itself while you do something else.

Please work through it **before** the Exercise 7 practical rather than during it. If you hit
a real problem, bring it to the start of the session and we'll help you, but we can't get
everybody set up and still have time to teach anything, so do try first.

You only ever do this once per computer. Unlike the Git and GitHub setup, there's nothing
here that expires, so you can do it as early in the course as you like.

\

## Step 1: Check you have the R markdown packages

Two packages do the work, `rmarkdown` and `knitr`. Both normally arrive with RStudio, so
check what you already have before installing anything. In the console:

```{r step1a, eval=FALSE}
library(rmarkdown)
library(knitr)
```

If neither of those gives you an error then you already have what you need and you can go
straight to Step 2. If either one complains that there is no such package, install it with:

```{r step1b, eval=FALSE}
install.packages("rmarkdown")
```

which brings `knitr` along with it.

[Appendix A](https://intro2r.com/install-rm.html) of the Introduction to R book covers this
in more detail, with separate instructions for Windows, macOS and Linux.

\

## Step 2: Install LaTeX

This is the step that takes the time, so start it and go and make a cup of tea.

To turn your document into a PDF, R markdown hands it to a typesetting system called LaTeX,
and you need one of those installed. The good news is that you don't have to learn anything
about it. You will never see it, you will never edit it, and once it's installed it simply
works in the background.

The easy way is a small LaTeX distribution called TinyTeX, which was built for exactly this
job. A full LaTeX installation is several gigabytes and needs its own installer, whereas
TinyTeX is a few hundred megabytes and installs itself from inside R. Run these two lines
in the console:

```{r step2, eval=FALSE}
install.packages("tinytex")

tinytex::install_tinytex()
```

The second line is the one that does the downloading. Give it a few minutes and a decent
connection, and don't be alarmed if it prints a great deal of text at you while it works.
You may also get a security or firewall prompt, which you should allow. Restart RStudio
once it has finished.

If you already have a LaTeX installation such as MiKTeX or MacTeX, you don't need TinyTeX
and can skip this step.

\

## Step 3: Check it worked

First, ask R:

```{r step3a, eval=FALSE}
tinytex::is_tinytex()
```

which should return `TRUE`.

That only tells you the installation is there, though. The real test is making a PDF, so do
that now rather than finding out in the session:

1. In RStudio, go to `File` -> `New File` -> `R Markdown...`
2. Put anything you like in the Title box, select **PDF** on the left, and click OK
3. RStudio opens a document already full of example content about cars. Leave it exactly as
   it is
4. Save it somewhere sensible, then click the **Knit** button at the top of the editor

If a PDF appears with a small table of car speeds and a plot in it, you're done and you can
delete the file.

\

## If something goes wrong

**The first knit fails with an error mentioning a file ending in `.sty`.** This is the
common one and it usually fixes itself. TinyTeX downloads the extra pieces it needs the
first time it meets them, so knit a second time before you do anything else.

**`is_tinytex()` returns `FALSE` after you installed it.** Restart RStudio and try again.
R needs restarting before it notices.

**You already had LaTeX installed and things are behaving oddly.** Having two can confuse
matters. Run `tinytex::tinytex_root()` to see which installation R is actually using.

**Knitting to HTML works but PDF doesn't.** That's the expected symptom of Step 2 not having
worked, since HTML needs no LaTeX at all. Go back and run it again.

If none of that helps, bring the error message with you to the session. Copy the whole
thing, not just the last line, because the useful part is usually further up.

\

## Where to go if you're stuck

The TinyTeX website at [yihui.org/tinytex](https://yihui.org/tinytex/) covers installation
problems in far more detail than this page does.

For R markdown itself, [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
is free online and is the standard reference.
